 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-06-22 paper.Rmd
update background section; correct case study 3

## Dimensionality Reduction Overview

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find low-dimensional representation of the data, $Y = [\mathbf{y_1},
\dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$ matrix where $d \ll p$.
The hope of the analyst is that the DR procedure to produces $Y$ will remove
noise in the original dataset while retaining latent structure.

DR methods can be classified into two broad groups: linear and non-linear
methods. Linear methods perform a linear transformation of the data, that is,
$Y$ is a linear transformation of $X$; one example is principal components
analysis (PCA) which performs an eigendecomposition of the estimated sample
covariance matrix. The eigenvalues are sorted in decreasing order and represent
the variance explained by each component (eigenvector). A common approach to
deciding on the number of principal components to retain is to plot the
proportion of variance explained by each component and choose a cut-off.

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Here we restrict our attention to two recent methods that are commonly used in
the literature: t-distributed stochastic neighbour embedding (t-SNE) and
uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
the pair-wise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

@@

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

The resulting configuration $Y$ is the one that minimizes the Kullback-Leibler
divergence between the probability distributions formed via similarities of
observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
$\mathcal{Q}$:

@@

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pair-wise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pair-wise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

@@

synthesis of some other techniques like LLE, Isomap and PHATE

@@

Like, when using DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of orthormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another. The tour is then the continous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases. A *grand tour* corresponds to choosing new orthoromal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections. Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function. While `liminal` is able to visualise guided tours, our focus in the case studies uses the grand tour to provide overviews of global
structure in the data.

@@

In the final case study, we explore a subset of the Google QuickDraw dataset [@quickdraw]. The data are sketches drawn by people of common concepts
like 'airplane' or 'shoe', and corresponding metadata about the humans who
created the sketches, such as country, the word category, whether the sketch
was recognised by Google's neural network, and the number of strokes taken.
We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R. We sampled 500 images from each of the following  animal categories: "dog", "cat", "octopus", "spider", and "whale" and unravelled the bitmaps into a 784 dimension image vector.

>>>>

## Dimensionality Reduction Background

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. Here we restrict our attention to two recent methods that are commonly used in the literature: t-distributed stochastic neighbour embedding 
(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
mention several other techniques.

To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produces $Y$ will remove noise in the original dataset while retaining any 
latent structure.

DR methods can be classified into two broad classes: linear and non-linear
methods. Linear methods perform a linear transformation of the data, that is,
$Y$ is a linear transformation of $X$; one example is principal components
analysis (PCA) which performs an eigendecomposition of the estimated sample
covariance matrix. The eigenvalues are sorted in decreasing order and represent
the variance explained by each component (eigenvector). 
A common approach to deciding on the number of principal components to retain is
to plot the proportion of variance explained by each component and choose a 
cut-off.

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct an embedding 
$Y$ such that the pariwise distances (inner products) in $Y$ approximate the pairwise distances (inner products) in  $X$. There are many variants of MDS, such as 
nonmetric scaling which uses ranks instead of distances. A related extension
of MDS is Isomap which uses the $k$-nearest neighbor graphs to estimate the pairwise geodesic distance of points in $X$ then uses classical MDS to construct $Y$. Yet
another approach are non-linear DR methods based on diffusion 
processes. A recent example in the single cell genomics literature is PHATE.
Here an affinity matrix is estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. The algorithm denoises estimated distances in 
high-dimensional space via transforming the affinity matrix into a Markov transition probaility matrix and diffusing this matrix over a fixed number of time steps. Then the diffused probabilities are transformed once more to construct a distance matrix, and classical MDS is used to generate $Y$. A difficulty with using non-linear
DR methods for exploratory data analysis are the choices required in selecting
appropriate parameters. To see the magnitude of these choices 
we now closely examine the precise details of t-SNE and UMAP.

The t-SNE algorithm estimates
the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pairwise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

@@

Next, the target low-dimensional space, $Y$, pairwise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

The resulting embedding $Y$ is the one that minimizes the Kullback-Leibler
divergence between the probability distributions formed via similarities of
observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
$\mathcal{Q}$:

@@

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pairwise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pairwise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

@@

@@

Like, when using DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) $A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the sense of continuous movement from one basis to another. The tour is then the continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the projection of $X$ onto $A_{t}$ as the tour path is interpolated between successive bases. A *grand tour* corresponds to choosing new orthonormal bases at random; allowing a user to ascertain structure via exploring the subspace of $d$-dimensional projections. Instead of picking projections at random, a *guided tour* can be used to generate a sequence 'interesting' projections as quantified by an index function. While `liminal` is able to visualise guided tours, our focus in the case studies uses the grand tour to provide overviews of global
structure in the data.

@@

In the final case study, we explore a subset of the Google QuickDraw dataset [@quickdraw]. The data are sketches drawn by people of common concepts
like 'airplane' or 'shoe', and corresponding metadata about the humans who
created the sketches, such as country, the word category, whether the sketch
was recognised by Google's neural network, and the number of strokes taken.
We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the `quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into a 784 dimension image vector. We sampled 500 images from each of the following  categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple", and "whale".
